---
title: "Table1"
author: "Travis Keep (Advisor: Dr. Seals)"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: Capstone Projects in Data Science
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

```{r, warning=FALSE, echo=T, message=FALSE}
library(betareg)
library(dplyr)
library(magrittr)
data("ReadingSkills")

dyslexia <- ReadingSkills %>% filter(dyslexia == "yes")
normal <- ReadingSkills %>% filter(dyslexia == "no")
n <- (ReadingSkills %>% dim)[1]
ndyslexia <- (dyslexia %>% dim)[1]
nnormal <- (normal %>% dim)[1]
iq_stats <- ReadingSkills %>% summarize(min(iq), median(iq), max(iq))
iq_dyslexia_stats <- dyslexia %>% summarize(min(iq), median(iq), max(iq))
iq_normal_stats <- normal %>% summarize(min(iq), median(iq), max(iq))
score_stats <- ReadingSkills %>% summarize(min(accuracy1), median(accuracy1), max(accuracy1))
score_dyslexia_stats <- dyslexia %>% summarize(min(accuracy1), median(accuracy1), max(accuracy1))
score_normal_stats <- normal %>% summarize(min(accuracy1), median(accuracy1), max(accuracy1))
```

Table1: Student ReadingSkills data stratified by dyslexia

| student characteristic | Total(n=`r n`) | Dyslexia(n=`r ndyslexia`) | Normal(n=`r nnormal`) |
|------------------------|-------------|----------------|--------------|
| IQ (Z score) | min `r iq_stats[1, 1]`<br>median `r iq_stats[1, 2]`<br>max `r iq_stats[1, 3]` |  min `r iq_dyslexia_stats[1, 1]`<br>median `r iq_dyslexia_stats[1, 2]`<br>max `r iq_dyslexia_stats[1, 3]` | min `r iq_normal_stats[1, 1]`<br>median `r iq_normal_stats[1, 2]`<br>max `r iq_normal_stats[1, 3]` |
| Reading Score | min `r score_stats[1, 1]`<br>median `r score_stats[1, 2]`<br>max `r score_stats[1, 3]` | min `r score_dyslexia_stats[1, 1]`<br>median `r score_dyslexia_stats[1, 2]`<br>max `r score_dyslexia_stats[1, 3]` |  min `r score_normal_stats[1, 1]`<br>median `r score_normal_stats[1, 2]`<br>max `r score_normal_stats[1, 3]` |


```{r, warning=FALSE, echo=T, message=FALSE}
# loading packages 
library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)
```

